---
title: "Project 1: NHL API Access and Analysis"
author: "Lee Pixton"
date: "6/20/2021"
output:
  pdf_document:
    toc: yes
    toc_depth: '4'
  rmarkdown::github_document:
    toc: yes
---

```{r setup_project, echo=F, warning=F, message=F}
library(httr)
library(knitr)
library(RSQLite)
library(jsonlite)
library(rmarkdown)
library(tidyverse)
```

# NHL API Connection Functions

This is the first project for the course ST558 (Data Science for Statisticians) at NC State University. In this vignette, we will discuss reading data from the National Hockey League (NHL) API, and then providing summaries of the data pulled.

## Required Project Packages 
In order to run the code for the project, the required packages, as well as code to install them, are below.

* httr
* knitr
* RSQlite
* jsonlite 
* tidyverse
* rmarkdown

### Installing required packages
To install the packages required for this project, use the code chunk below, removing the beginning `#`
```{r install}
#install.packages("knitr", "httr", "RSQlite", "jsonlite", "rmarkdown", "tidyverse")
```


## Function to Contact Franchise API
First thing to do is create a function that can access the franchise API. This will allow us to pull any number of information about each franchise. The `extension` option in the function takes in one of the following:

* `franchise` - Returns id, firstSeasonId and lastSeasonId and name of every team in the history of the nhl
* `franchise-team-totals` - Returns total stats for every franchise (ex roadTies, roadWins, etc)
* `franchise-season-records` - Returns drill-down into season records for a specific franchise
* `franchise-goalie-records` - Returns goalie records for the specified franchise
* `franchise-skater-records` - Returns skater records, same interaction as goalie endpoint
* `franchise-detail` - Returns captainHistory, coachingHistory, generalManagerHistory and a summary of retired numbers


And the `ID` option is available for specifying a team.

```{r franchise function, warning=F}
# This function is used to take input from the user and return appropriate data from the NHL franchise API
franchiseAPI <- function(extension, ID = NULL){
  base <- "https://records.nhl.com/site/api/"
  
  if(is.null(extension)){
    return("Must include a valid franchise call")
  } else{
    if(!is.null(ID)){
      if(extension == "franchise" | extension == "franchise-team-totals"){
        warning("ID is not allowed for these calls, will return ", extension, "without ID")
        URL <- paste0(base, extension)
      } else if(extension == "franchise-detail"){
        URL <- paste0(base, extension, "?cayenneExp=mostRecentTeamId=", ID) 
      } else{
        URL <- paste0(base, extension, "?cayenneExp=franchiseId=", ID)
      }
    } else{
      URL <- paste0(base, extension) 
    }
  }
  
  get_nhl <- GET(URL)
  nhl_text <- content(get_nhl, "text")
  nhl_json <- fromJSON(nhl_text, flatten=T)
  return(tbl_df(nhl_json$data))
}
```

### ID Mapping Table
Once we have the function, the next step is to put together a mapping from team name to ID number. This allows the user to input the team name without having to know the specific ID. We will use our API call to retrieve this information into a dataframe, that we can then use to complete a mapping in the wrapper function. The user will be able to use the full name, the abbreviation, or just the mascot to pull the ID.
```{r team mapping, warning=F,message=F}
team_mapping <- franchiseAPI("franchise")
team_mapping <- team_mapping %>% select(!(ends_with("SeasonId") | id | ends_with("PlaceName")))
kable(team_mapping)
```


## Function to Contact Stats API
Next we need a function to access the stats API. Though there are many modifiers available to access multiple data sets, for our purposes we will only provide access to the `?expand=team.stats` modifier. By providing an `ID`, it will pull specific team data, but the function alone will pull data on all the teams.

```{r stats function, warning=F}
# This function is used to take user input and output a tibble with data from the NHL Stats API
statsAPI <- function(ID = NULL){
  base <- "https://statsapi.web.nhl.com/api/v1/teams?expand=team.stats"
  if(!is.null(ID)){
    URL <- paste0(base,"=",ID)
  } else{
    URL <- paste0(base)
  }
  
  get_stats <- GET(URL)
  stats_text <- content(get_stats, "text")
  stats_json <- tbl_df(fromJSON(stats_text, flatten=T))
  stats_data <- stats_json$teams$teamStats
  
  if(!is.null(ID)){
    stats_data <- stats_json$teams$teamStats[stats_json$teams$id == ID]
    end_data <- stats_data[[1]]$splits[[1]]
  } else{
    end_data <- stats_data[[1]]$splits[[1]]
    
    for(i in c(2:length(stats_data))){
      stats_data <- stats_json$teams$teamStats
      end_data <- rbind(end_data,stats_data[[i]]$splits[[1]])
    }
  }
  return(tbl_df(end_data))
}
```

## Wrapper Function for Both API Calls
Here we create a function that takes in the user input specific to either API.
```{r wrapper_function}
# Allows for one function call to query either API
access_NHL_API <- function(extension, ID = NULL, ...){
  if(!is.numeric(ID) & !is.null(ID)){
    if(length(which(team_mapping$fullName == ID)) == 0){
      i <- which(team_mapping$fullName == ID)
    } else if(length(which(team_mapping$teamAbbrev == ID)) == 0){
      i <- which(team_mapping$teamAbbrev == ID)
    } else if(length(which(team_mapping$teamCommonName == ID)) == 0){
      i <- which(team_mapping$teamAbbrev == ID)
    } else{
      stop("Team ID not found!")
    }
  }
  
  if (extension == 'stats'){
    if(!is.null(ID)){
      ID <- team_mapping$teamAbbrev[i]
    }
    return(statsAPI(ID))
  } else {
    if(!is.null(ID)){
      ID = i
    }
    return(franchiseAPI(extension,ID))
  }
}
```

# NHL Data Analysis Using Our API
First we need to pull the data. We will pull franchise data, team totals, and current season stats. We will conduct some filtering here to make the join simpler.
```{r pull data, warning=F,message=F}
franchise <- access_NHL_API("franchise") %>% select(!(lastSeasonId | id))
team_totals <- access_NHL_API("franchise-team-totals") %>% filter(gameTypeId == 2 & activeFranchise == 1 & is.na(lastSeasonId)) %>% select(!(ends_with("SeasonId") | id))
stats <- access_NHL_API("stats") %>% filter(!is.na(stat.gamesPlayed))
```

## Combining Data
Next we need to combine these data together. 
```{r combine data}
final_data <- right_join(franchise, team_totals, by = c("mostRecentTeamId" = "teamId")) %>% inner_join(stats, by = c("teamName" = "team.name")) 
```

## Adding Columns
In this section we will do some calculations to add a couple columns of interest. We will then filter on columns of interest. Six of the final columns are shown below.

* `seasonsInLeague` - How many seasons the team has been in the league
* `currentWinPerc` - Win percentage of the current season
* `currentGoalDiff` - Goal differential (team goals scored - team goals allowed) for current season
* `histWinPerc` - Historic win percentage
* `histGoalsPerGame` - Historic goals per game
* `histGoalDiff` - Historic goal differential
* `winPercDiff` - Difference between current season win percentage and historic
* `seasonGoalDiff` - Difference between current goal differential and historic
* `goalsPerGameDiff` - The difference between the current season goals per game and historic

```{r add col}
final_data$firstSeason <- as.integer(substr(as.character(final_data$firstSeasonId),1,4))
final_data <- final_data %>%
  mutate(
    seasonsInLeague = 2021 - firstSeason,
    currentWinPerc = as.integer(stat.wins) / stat.gamesPlayed,
    currentGoalDiff = as.numeric(stat.goalsPerGame) - as.numeric(stat.goalsAgainstPerGame),
    histWinPerc = wins / gamesPlayed,
    histGoalsPerGame = goalsFor / gamesPlayed,
    histGoalDiff = histGoalsPerGame - (goalsAgainst / gamesPlayed),
    winPercDiff = currentWinPerc - histWinPerc,
    seasonGoalDiff = currentGoalDiff - histGoalDiff,
    goalsPerGameDiff = as.numeric(stat.goalsPerGame) - histGoalsPerGame
  ) %>% 
  select(fullName,teamAbbrev,seasonsInLeague, gamesPlayed, goalsFor, goalsAgainst, histGoalsPerGame, histGoalDiff, wins, losses, histWinPerc, stat.wins, stat.losses, currentWinPerc, stat.goalsPerGame, stat.goalsAgainstPerGame, currentGoalDiff, winPercDiff, seasonGoalDiff, goalsPerGameDiff)
kable(final_data[,1:6])
```

## Exploratory Data Analysis
Now that we have our data set with 31 rows and 20 columns, we are able to do exploratory data analysis (EDA).
```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```





